A forest is more than its trees: haplotypes and ancestral recombination graphs

### 2024-12-18 paper/writeup.tex
one more citation

\paragraph{IBD in ARGs}
The term ``identity by descent'' (IBD) is used to mean many different (but related) things,
and length distributions of shared IBD segments can be used for inference of recent demographic history
\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate,yang2016rapidly,silcocks2023indigenous,huang2024estimating}.
A commonly-used definition in the context of a given ARG says that
the two genomes share an IBD segment if each has inherited the segment from their common ancestor
along a single path \citep{ralph2013geography}.
Largely for computational reasons, this is the definition that is used in \tskit's
IBD-finding methods 
(by G.~Tsambos \citet{tskit}; see also \citet{tsambos2023link}).
However, many simulation and inference methods produce ARGs
as shown in \figref[A\&B]{fig:conceptual},
in which inheritance of a single segment is represented by more than one edge.
This means that the \texttt{ibd\_segments} method of \tskit{} will return shorter segments than it ought to.
However, applying our method of extending haplotypes
will remove this issue, and so should make returned IBD segments
more closely match those predicted by theory.
Further work is needed to understand how accurately IBD segments
are inferred by various ARG inference methods.

>>>>

\paragraph{IBD in ARGs}
The term ``identity by descent'' (IBD) is used to mean many different (but related) things,
and length distributions of shared IBD segments can be used for inference of recent demographic history
\citep[for instance,][]{alasadi2018estimating,ringbauer2017inferring,browning2015accurate,yang2016rapidly,silcocks2023indigenous}.
A commonly-used definition in the context of a given ARG says that
the two genomes share an IBD segment if each has inherited the segment from their common ancestor
along a single path \citep{ralph2013geography}.
Largely for computational reasons, this is the definition that is used in \tskit's
IBD-finding methods 
(by G.~Tsambos \citet{tskit}; see also \citet{tsambos2023link}).
However, many simulation and inference methods produce ARGs
as shown in \figref[A\&B]{fig:conceptual},
in which inheritance of a single segment is represented by more than one edge.
This means that the \texttt{ibd\_segments} method of \tskit{} will return shorter segments than it ought to.
However, as we have shown above,
our method of extending haplotypes will modify the tree sequence
so that the inherited segment is represented by a single edge
(as in \figref[C\&D]{fig:conceptual}).
So, if our method (\texttt{extend\_haplotypes}) is applied before finding IBD segments
(with  \texttt{ibd\_segments}),
then the resulting segments should much more accurately represent
the IBD segments (in the ``path'' sense used here) implied by the tree sequence.
Whether the resulting segments better match those predicted by theory depends
also on the quality of tree inference.
Note also that \citet{huang2024estimating} and \citet{guo2024strong}
both provide methods for tree sequences to compute a different definition of IBD
(segments on which the MRCA does not change),
which is unaffected by this issue.
Further work is needed to understand how accurately IBD segments
are inferred by various ARG inference methods.
